---
title: "East Link Fall 2024, Spring 2025, & Ongoing Fall 2025"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float:
      collaspe: false
      smooth_scroll: true
      fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# github location
#source(here::here('analyses', 'kcm_style.R'))
#MZ attempted to fix source of R code 
source(here::here('ELC monitoring', 'ELC-monitoring', 'elink_MZedits.R'))



library(kableExtra)
library(plotly)
library(sf) # for shapes
library(leaflet) # for interactive maps
library(leafsync)


# Render Command
# rmarkdown::render(input = here::here('analyses', 'rtw',"rtw_rmd.Rmd"), output_format = "html_document", envir = globalenv())

```

## Summary

-   Weekday stop ridership (ons + offs)
-   Examines Fall 2024, Spring 2025, and Fall 2025 data (does NOT include DART route currently):
    -   In total, there are `r length(route_list)` routes and `r length(stop_list)` stops: `r route_list`
    -   In Fall 24, there are `r count_route_243` routes and `r count_stop_243` stops: `r route_243`
    -   In Spring 25, there are `r count_route_251` routes and `r count_stop_251` stops: `r route_251`
    -   In Fall 25, there are `r count_route_253` routes and `r count_stop_253` stops: `r route_253`

## Stop Ridership

### Spring 2025 - Fall 2025

Note: there is no Fall 2025 dart data yet

```{r}

kbl(
  summ_table(
    source = 'stop_ridership',
    group_var = c('Day', 'dart'),
    #select_service_change = c(243, 251),
    select_service_change = c(251, 253),
    var1 = c('rider')
  ) %>%
    # MZ commented out to get all days of week: filter(Day == 'Weekday') %>%
    #janitor::adorn_totals() %>%
    #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
    mutate(
      # Change = get('Spring 2025') - get('Fall 2024'),
      # Percent_Change = Change / get('Fall 2024')
      Change = get('Fall 2025') - get('Spring 2025'),
      Percent_Change = Change / get('Spring 2025')
    ) %>%
    # MZ updated Percent_Change accuracy to one decimal place 0.1
    mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 0.1)) %>%
    mutate_if(is.numeric, scales::label_comma()) %>%
    select(-Variable)
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

#MZ attempt to define another color palette with 5 colors and 5 even breaks
pal <- colorFactor("Spectral",data$colors, n=5)

map_stop_change <- stop_ridership_map(
  # before = 243,
  # after = 251,
  before = 251,
  after = 253,
  var1 = 'average daily ridership',
  peak = 'no',
  change = 'yes'
  # MZ attempt to change colors
  fillColor = ~ pal(colors())

)

#leafsync::sync(map_stop_change1, map_stop_change)

map_stop_change

map_stop_253 <- stop_ridership_map(
  # before = 243,
  # after = 251,
  before = 251,
  after = 253,
  var1 = 'average daily ridership',
  peak = 'no',
  change = 'no'
)
map_stop_253

```

### By Time of Day

Weekday data only Note: dart data are not available by period

```{r}

ridership_crosstab_overtime_plt(
  source = 'stop_ridership',
  day_of_week = 'Weekday',
  select_service_change = service_changes,
  x_axis = 'time_period_at_stop',
  var1 = 'rider'
)


kbl(
  summ_table(
    source = 'stop_ridership',
    group_var = c('Day', 'time_period_at_stop'),
    #select_service_change = c(243, 251),
    select_service_change = c(251, 253),
    var1 = c('rider')
  ) %>%
    filter(Day == 'Weekday') %>%
    #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
    mutate(
      Change = get('Fall 2025') - get('Spring 2025'),
      Percent_Change = Change / get('Spring 2025')
      # Change = get('Spring 2025') - get('Fall 2024'),
      # Percent_Change = Change / get('Fall 2024')
    ) %>%
    arrange(desc(Percent_Change)) %>%
    mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
    janitor::adorn_totals() %>%
    mutate_if(is.numeric, scales::label_comma()) %>%
    filter(!is.na(time_period_at_stop)) %>%
    select(-Variable, -Day)
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### By Route

Note: stop data are summed by route, however for route level ridership comparisons, use trip level data

```{r}

ridership_crosstab_overtime_plt(
  source = 'stop_ridership',
  day_of_week = 'Weekday',
  select_service_change = c(243, 251, 253),
  x_axis = 'hour',
  var1 = 'rider'
)


ridership_crosstab_overtime_plt(
  source = 'stop_ridership',
  day_of_week = 'Weekday',
  select_service_change = c(243, 251, 253),
  x_axis = 'route',
  var1 = 'rider'
)

kbl(
  summ_table(
    source = 'stop_ridership',
    group_var = c('Day', 'route', 'dart'),
    #select_service_change = c(243, 251),
    select_service_change = c(251, 253),
    var1 = c('rider')
  ) %>%
    filter(Day == 'Weekday') %>%
    #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
    mutate(
      Change = get('Fall 2025') - get('Spring 2025'),
      Percent_Change = Change / get('Spring 2025')
      # Change = get('Spring 2025') - get('Fall 2024'),
      # Percent_Change = Change / get('Fall 2024')
    ) %>%
    arrange(desc(Percent_Change)) %>%
    mutate(
      Percent_Change = scales::percent(Percent_Change, accuracy = 1),
      route = as.character(route)
    ) %>%
    janitor::adorn_totals() %>%
    mutate_if(is.numeric, scales::label_comma()) %>%
    select(-Variable, -Day)
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Trip Ridership

Overall by day

```{r}

# kbl(
#   summ_table(
#     source = 'trip_ridership',
#     group_var = c('Day'),
#     select_service_change = c(243, 251),
#     var1 = c('ons')
#   ) %>%
#     #filter(Day == 'Weekday') %>%
#     #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
#     mutate(
#       Change = get('Spring 2025') - get('Fall 2024'),
#       Percent_Change = Change / get('Fall 2024')
#     ) %>%
#     arrange(desc(Percent_Change)) %>%
#     mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
#     janitor::adorn_totals() %>%
#     mutate_if(is.numeric, scales::label_comma()) %>%
#     select(-Variable)
# ) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

# kbl(
#   summ_table(
#     source = 'trip_ridership',
#     group_var = c('Day'),
#     select_service_change = c(251, 253),
#     var1 = c('ons')
#   ) %>%
#     #filter(Day == 'Weekday') %>%
#     #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
#     mutate(
#       Change = get('Fall 2025') - get('Spring 2025'),
#       Percent_Change = Change / get('Spring 2025')
#     ) %>%
#     arrange(desc(Percent_Change)) %>%
#     mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
#     janitor::adorn_totals() %>%
#     mutate_if(is.numeric, scales::label_comma()) %>%
#     select(-Variable)
# ) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

```

Weekdays

```{r}
#weekdays
# kbl(
#   summ_table(
#     source = 'trip_ridership',
#     group_var = c('Day', 'route'),
#     select_service_change = c(243, 251),
#     var1 = c('ons')
#   ) %>%
#     filter(Day == 'Weekday') %>%
#     #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
#     mutate(
#       Change = get('Spring 2025') - get('Fall 2024'),
#       Percent_Change = Change / get('Fall 2024')
#     ) %>%
#     arrange(desc(Percent_Change)) %>%
#     mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
#     janitor::adorn_totals() %>%
#     mutate_if(is.numeric, scales::label_comma()) %>%
#     select(-Variable, -Day)
# ) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

# ridership_route_service_change_plt(
#   day_of_week = 'Weekday',
#   #day_of_week = c('Weekday', 'Saturday', 'Sunday'),
#   select_routes = route_list,
#   select_service_change = service_changes,
#   var1 = 'ons'
# )
# #route_ridership_map()
```

weekends

```{r}
# weekend
# ridership_route_service_change_plt(day_of_week = c('Saturday', 'Sunday'))
# ridership_route_service_change_plt(select_service_change = c(193,251),
#                                    day_of_week = c('Weekday', 'Saturday', 'Sunday'))
```

Saturday

```{r}

# kbl(
#   summ_table(
#     source = 'trip_ridership',
#     group_var = c('Day', 'route'),
#     select_service_change = c(243, 251),
#     var1 = c('ons')
#   ) %>%
#     filter(Day == 'Saturday') %>%
#     #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
#     mutate(
#       Change = get('Spring 2025') - get('Fall 2024'),
#       Percent_Change = Change / get('Fall 2024')
#     ) %>%
#     arrange(desc(Percent_Change)) %>%
#     mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
#     janitor::adorn_totals() %>%
#     mutate_if(is.numeric, scales::label_comma()) %>%
#     select(-Variable, -Day)
# ) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))
```

Sunday

```{r}

# kbl(
#   summ_table(
#     source = 'trip_ridership',
#     group_var = c('Day', 'route'),
#     select_service_change = c(243, 251),
#     var1 = c('ons')
#   ) %>%
#     filter(Day == 'Sunday') %>%
#     #mutate(Change = .[ncol(.)] - .[ncol(.) -1]) %>%
#     mutate(
#       Change = get('Spring 2025') - get('Fall 2024'),
#       Percent_Change = Change / get('Fall 2024')
#     ) %>%
#     arrange(desc(Percent_Change)) %>%
#     mutate(Percent_Change = scales::percent(Percent_Change, accuracy = 1)) %>%
#     janitor::adorn_totals() %>%
#     mutate_if(is.numeric, scales::label_comma()) %>%
#     select(-Variable, -Day)
# ) %>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

```
